---
title: "Analysis and quantification of mass spectrometry imaging data"
author: "Dr Matt Smith"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    html_document :
      number_sections: false
      toc: true
      toc_float: 
        collapsed: false
        smooth_scroll: true
---


```{r setup, include=FALSE}

# set default options
knitr::opts_chunk$set(echo = FALSE,
               message = FALSE,
               warning = FALSE,
               fig.align="center",
               fig.width = 5,
               fig.height = 5,
               dpi = 120,
               fig.retina = 3)

```

# Load packages and functions
```{r, message=FALSE}
library(quantMSImageR)
library(Cardinal)
library(dplyr)
library(chemCal)
library(openxlsx)

# Set paths
base_path = "C:/Users/matsmi/OneDrive - Karolinska Institutet/Dokument/MSI/quantMSImageR"
data_path = "C:/Users/matsmi/OneDrive - Karolinska Institutet/Dokument/MSI/quantMSImageR/data" 
out_path = "C:/Users/matsmi/OneDrive - Karolinska Institutet/Dokument/MSI/quantMSImageR/results"

lib_ion_path = sprintf("%s/ion_library.txt", data_path)

fns = c("19Jan_lung_01_LHS", "19Jan_lung_01_RHS",
        "21Jan_lung_01-LHS", "21Jan_lung_01-RHS", "21Jan_lung_02-LHS",
        "22Jan_lung_01-LHS", "22Jan_lung_01-RHS", "22Jan_lung_02-LHS_Acq01", "22Jan_lung_02-RHS_Acq02", "22Jan_lung_02-LHS_Acq01_rep")

fn = fns[10]

```

# Load files

Read imzML mass spectrometry imaging data files into R, using Cardinal package and generate ion Cardinal::images to view the data.
Optionally, write to imzML if necessary to use with existing workflows for instance.

```{r, message=FALSE}

tissue = read_mrm(name = fn, folder = data_path, lib_ion_path = lib_ion_path , polarity = "Negative")

#Cardinal::writeImzML(tissue, name=fn, folder=out_path, merge=T, mz.type="32-bit float", intensity.type="32-bit float")

```

# Set regions of interest (ROIs)

Select ROIs in each MSI dataset.

## Subset tissue

Select ROI around the tissue of interest from the MSI dataset and optinally save Cardinal::image for future use.


```{r, message=FALSE, fig.align='center', fig.show='hold', out.width="49%",out.height="49%"}
#Cardinal::image(tissue, feature = 2)

# Subset tissue pixels
tissue_pixels = selectROI(tissue, contrast.enhance="histogram", feature= 3)
tissue = tissue[, tissue_pixels]

# Save and load Cardinal::image
save.image(file = sprintf("%s/rdata/%s.RData", out_path, fn))
load(file = sprintf("%s/rdata/%s.RData", out_path, fn))


```

## Investigate data {.tabset}

### Feature metadata  {.tabset .tabset-fade}

```{r}

# Check feature metadata
DT::datatable(data.frame(fData(tissue)))
write.csv(data.frame(fData(tissue)), sprintf("%s/fdata/%s.csv", out_path, fn), row.names=F)

```


Scroll ion images

```{r}

#for(i in 1:nrow(tissue)){
  
#  comp = fData(tissue)$name[i]
  
#  cat( paste('#### ', comp, '\n' ) )
  
#  Cardinal::image(tissue, contrast.enhance="histogram", smooth.image = 'gaussian', superpose = FALSE, normalize.image = "linear", feature =fData(tissue)@mz[i], main=comp, adj=0, line=1)
  
#  Cardinal::image(tissue, contrast.enhance="suppression", smooth.image = 'gaussian', superpose = FALSE, normalize.image = "linear", feature =fData(tissue)@mz[i], main=comp, adj=0, line=1)
  
#  cat('\n \n')
#}

############ Plot (histogram enhance)
png(sprintf("%s/ion_images/%s_histogram.png", out_path, fn), width = 9, height = 7, units = "in", res = 1200)
plot.new()
Cardinal::image(tissue, contrast.enhance="histogram", smooth.image = 'gaussian', superpose = FALSE, normalize.image = "linear", feature =fData(tissue)@mz[1:nrow(fData(tissue))])
dev.off()


############ Plot (suppression enhance)
png(sprintf("%s/ion_images/%s_suppression.png", out_path, fn), width = 9, height = 7, units = "in", res = 1200)
plot.new()
Cardinal::image(tissue, contrast.enhance="suppression", smooth.image = 'gaussian', superpose = FALSE, normalize.image = "linear", feature = fData(tissue)@mz[1:nrow(fData(tissue))])
dev.off()

```


## Multivariate stats

```{r}

#dgmm <- spatialDGMM(tissue, r=20, k=2, method="adaptive")
#dgmm <- spatialDGMM(tissue[2,], r=20, k=3, method="adaptive")

#summary(dgmm)

#Cardinal::image(dgmm)

#plot(dgmm)

# Extract the average amount per pixel at each ROI in tissue
pData(tissue)$sample_ID = 1:nrow(pData(tissue))
tissue_df <- createMSIDatamatrix(MSIobject = as(tissue, "quant_MSImagingExperiment"), roi_head = NA)


conc_matrix = tissue_df@tissueInfo@conc_matrix
sample_metadata = tissue_df@tissueInfo@sample_metadata
feature_metadata = data.frame(fData(tissue_df))

conc_matrix
head(sample_metadata)
head(feature_metadata)

# Save xlsx

output_matrix = createWorkbook()
# variable metadata
addWorksheet(output_matrix, "feature_metadata")
writeData(output_matrix, "feature_metadata", feature_metadata, colNames=TRUE, rowNames = FALSE, keepNA=FALSE)
# sample metadata
addWorksheet(output_matrix, "sample_metadata")
writeData(output_matrix, "sample_metadata", sample_metadata, colNames=TRUE, rowNames = FALSE, keepNA=FALSE)
# peak area matrix
addWorksheet(output_matrix, "peak_area_matrix")
writeData(output_matrix, "peak_area_matrix",
          t(conc_matrix),
          colNames=TRUE, rowNames = TRUE, keepNA=FALSE)
# save workbook
saveWorkbook(output_matrix,
             file = sprintf("%s/dataframe/%s.xlsx", out_path, fn),
             overwrite = TRUE)

```
